**Effective application of machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-09-11 submission/manuscript.Rmd
version sent to Jenna

######################################################################
#----------------- Load feature importance data -----------------#
# Median of base testing AUC for each model
# 100 datasplit permutation importances new testing AUC for top 5 features
######################################################################

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes, and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but inconsistent adoption of methods call the validity of these models into question. Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability. To overcome these challenges, we trained seven models that used fecal 16S rRNA sequences to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We developed a generalizable pipeline to train, validate and interpret the models. To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting SRNs with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, it trained much faster (12 min) and was more interpretable. Our analysis showed that ML models should be chosen based on the goal of the study as it will inform our expectations of predictive performance and interpretability.

@@

Prediction of health outcomes using machine learning (ML) is rapidly being adopted in human microbiome studies. However, these ML models are likely overoptimistic in terms of quantifying predictive performance. Moreover, there is a trend towards using black box models such as random forest and neural networks without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease. This work represents a step towards developing better ML practices in microbiome research by implementing a rigorous pipeline and emphasizing the importance of selecting ML models that reflect the goal of the study. These concepts are not particular to study of health outcomes but can also be applied to environmental microbiology studies as well.

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases. However, the structure of the human microbiome is remarkably variable between individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population individually. Recently, machine learning models have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017]. The task of diagnosing an individual with high confidence relies on a model that is built with rigorous methods. However, there are common methodological problems across many of these studies that need to be addressed as the field progresses. These include a lack of transparency in which methods are used and how these methods are implemented; developing and evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances. Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing analysis frameworks to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. More work is needed to further improve reproducibility and minimize over-optimism for model performance.

@@

__Model selection and pipeline construction__ We established a ML pipeline where we trained and validated each of the seven models using a common approach that is based on standard methods within the ML community (REFS by Jenna?)[Figure 1].

@@

Second, we trained seven different models using the training data [Table 1]. We selected models with different classification algorithms and regularization methods. Regularization is a technique that discourages overfitting by penalizing the model for learning the training data too well. For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features. For comparison, we also trained an L1-regularized SVM model with linear kernel. L1-regularization on microbiome data led to a sparser solution (i.e., force many coefficients to zero). To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including decision tree, random forest, and XGBoost and we trained an SVM model with non-linear kernel.

Third, fitting of these models require selecting appropriate hyperparameters. Hyperparameters are the rules that are learned from the training set in a classification algorithm. For example, in the linear models the regularization term (C) is a hyperparameter that indicates the penalty for overfitting. Similar to regularization term C, all hyperparameters are tuned to find the best model. We selected hyperparameters by performing 100 five-fold cross-validation (CV) repeats on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the best hyperparameter values for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model’s predictions are no different from random chance. To select the best performing hyperparameter, we performed a full grid search for hyperparameter settings when training our models. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new ML task. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.

Finally, we trained the full training dataset with the selected hyperparameter values and applied the model to the held-out data to evaluate the testing predictive performance of each model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of seven models to classify individuals as having normal colons or SRNs [Figure 2]. The random forest model had significantly higher test AUROC values than the other models for detecting SRNs (Wilcoxon rank sum test, p < 0.01). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and SVM with linear kernel) performed nearly as well as random forest.

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that would indicate that the models were overfit. The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]; however, these differences are relatively small and would not indicate a problem with overfitting.

To evaluate the risk for over-optimism of each model, we calculated the range of AUROC values for each model using 100 data-splits. The range among the testing AUROC values within each model varied by `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` on average across the seven models. If we had only done a single split, then there is a risk that we could gotten lucky or unlucky with the performance of the model. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`. These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2]. Therefore, it is important to employ the hierarchical data splits that were included in our pipeline to minimize the risk of over-optimism.

@@

__Interpretation of each ML model.__ Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017]. Because we often use ML models not just to predict a health outcome but also to learn the ecology behind a disease, model interpretation becomes crucial for microbiome studies. ML models decrease in interpretability as they increase in complexity. In this study we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3]. We also reviewed the signs of feature weights to determine whether an OTU is associated with classifying a subject as being healthy or having an SRN - negative sign indicated being healthy and positive sign indicated having an SRN. It was encouraging that many of the highest ranked OTUs were shared across these three models, (e.g. OTU 50, 426, 609, 822, 1239). The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU. However, this approach was only possible with linear models.

Second, to analyze non-linear models we interpreted the feature importance using permutation importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had a minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(round(RF$imp, 3))`. Effectively, the complexity of the communities was more fully represented in the better performing models [@sze_leveraging_2018; @wirbel_meta-analysis_2019]. At least in this case, it was not possible to distinguish between health and disease using a single OTU. While permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU is associated with classifying a subject as being healthy or having an SRN.

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S4]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S4], 17 out of the 20 top OTUs (e.g. OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. These results indicate that both methods are consistent in selecting the most important OTUs.

@@

There is a growing awareness that many human diseases and environmental processes are not driven by a single organism but are the product of multiple bacterial populations. Traditional statistical approaches are useful for identifying those cases where a single organism is associated with a process. In contrast, ML methods offer the ability to incorporate the structure of the microbial communities as a whole to classify them into different categories such as coming from a patient who is healthy or has SRNs. If it is possible to classify communities reliably, then ML methods also offer the ability to identify those microbial populations within the communities that are responsible for the classification. However, the application of ML in microbiome studies is still in its infancy and the field still needs to develop a better understanding of different ML methods, their strengths and weaknesses, and how to implement them.

To address these needs, we developed a framework to train rigorous, transparent, and reproducible models. We benchmarked seven ML models and showed that we can create models that are inherently interpretable and easily trained without losing predictive performance. In terms of predictive performance, the random forest model had the best testing AUROC values compared to the other six models. However, the second-best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest was a complex ML model and could only be explained using post-hoc methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights. Comparing many different models showed us that the most complex model is not necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome. We performed a random datasplit to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We repeated this datasplit 100 times to measure predictive performance. During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the similar AUROC values from the cross-validation and testing.

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures which create collinearity in the datasets we work with. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. For example if one OTU is highly correlated with another OTU, only one of these OTUs would have a large feature weight thus hiding the importance of its correlated OTU. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features. It is important to know the correlation strutures of the data to avoid missinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronunced due to many genes being correlated with one another because they come from the same chromosome. To identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.

In this study, we did not consider all possible modeling approaches. However, the principles highlighted throughout this study apply to all ML modeling tasks with microbiome data. For example, we did not evaluate multicategory classification methods to predict non-binary outcomes. For example, we could have trained models to differentiate between people with normal colons and those with adenomas or carcinomas (k=3 categories). We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs. Furthermore, as number of categories to classify increase, more samples are required for each category to train a model. We also did not use regression-based analyses to predict a non-categorical outcome. We have previously used such an approach to train random forest models to predict fecal short-chain fatty acid concentrations based on microbiome data [@sze_fecal_2019]. Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks. Deep learning methods hold great promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which makes the deep learning models prone to overfitting. These methods are even more complex than random forest and XGBoost and are considered uninterpretable. There is great potential for applying ML approaches to microbiome data and we can choose which ML approach is best for the study.

@@

\newpage
\captionof{table}{Characteristics of the machine learning models in our comparative study.}
\small
\begin{tabular}{|l|l|l|l|l|}
\hline

>>>>

######################################################################
#----------------- Load feature importance data -----------------#
# Median of base testing AUC for each model
# 100 data-split permutation importances new testing AUC for top 5 features
######################################################################

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training methods call the validity of these models into question. Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability. To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We developed a generalizable pipeline to train, validate and interpret the models. To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model performed best at detecting SRNs with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, trained much faster (12 min), and was more interpretable. Our analysis showed that ML models should be chosen based on the goal of the study as the choice will inform expectations of performance and interpretability.

@@

Prediction of health outcomes using machine learning (ML) is rapidly being adopted in human microbiome studies. However, these ML models are likely overoptimistic in terms of quantifying predictive performance. Moreover, there is a trend towards using black box models such as random forest and neural networks without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease. This work represents a step towards developing better ML practices in microbiome research by implementing a rigorous pipeline and emphasizing the importance of selecting ML models that reflect the goal of the study. These concepts are not particular to the study of health outcomes but can also be applied to environmental microbiology studies.

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases. However, the structure of the human microbiome is remarkably variable between individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population individually. Recently, machine learning models (ML) have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017]. The task of diagnosing an individual with high confidence relies on a model that is built with rigorous methods. However, there are common methodological problems across many of these studies that need to be addressed as the field progresses. These include a lack of transparency in which methods are used and how these methods are implemented; developing and evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances. Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. More work is needed to further improve reproducibility and minimize over-optimism for model performance.

@@

__Model selection and pipeline construction__ We established a common ML pipeline to train and validate each of the seven models that is based on standard methods within the ML community (REFS by Jenna?)[Figure 1].

@@

Second, we trained seven different models using the training data [Table 1]. We selected models with different classification algorithms and regularization methods. Regularization is a technique that discourages overfitting by penalizing the model for learning the training data too well. For regularized logistic regression and SVM with linear kernel, we used L2-regularization to keep all potentially important features. For comparison, we also trained an L1-regularized SVM model with linear kernel. L1-regularization on microbiome data led to a sparser solution (i.e., forced many coefficients to zero). To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including decision tree, random forest, and XGBoost and we trained an SVM model with a non-linear kernel.

Third, fitting of these models required selecting appropriate hyperparameters. Hyperparameters are the rules that are learned from the training set in a classification algorithm. For example, in the linear models the regularization term (C) is a hyperparameter that indicates the penalty for overfitting. Similar to regularization term C, all hyperparameters are tuned to find the best model. We selected hyperparameters by performing 100 five-fold cross-validation (CV) repeats on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the best hyperparameter values for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model’s predictions are no different from random chance. To select the best performing hyperparameter, we performed a full grid search for hyperparameter settings when training our models. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new ML task. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.

Finally, we trained the full training dataset with the selected hyperparameter values and applied the model to the held-out data to evaluate the testing predictive performance of each model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to obtain a reliable and robust interpretation of model performance [Figure 1].

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2]. The random forest model had significantly higher test AUROC values than the other models for detecting SRNs (Wilcoxon rank sum test, p < 0.01). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(format(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(format(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and SVM with linear kernel) performed nearly as well as random forest.

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that would indicate that the models were overfit. The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]. These differences are relatively small and would not indicate a problem with overfitting.

To evaluate the risk for over-optimism of each model, we calculated the range of AUROC values for each model using 100 data-splits. The range among the testing AUROC values within each model varied by `r paste(format(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3), nsmall=3))` on average across the seven models. If we had only done a single split, then there is a risk that we could gotten lucky or unlucky with the performance of the model. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(format(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3), nsmall=3))`. These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2]. Therefore, it is important to employ the 100 hierarchical data splits that were included in our pipeline to minimize the risk of over-optimism.

@@

__Interpretation of each ML model.__ Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017]. Because we often use ML models not just to predict a health outcome but also to identify the biomarkers for a disease, model interpretation becomes crucial for microbiome studies. ML models decrease in interpretability as they increase in complexity. In this study we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3]. We also reviewed the signs of feature weights to determine whether an OTU was associated with classifying a subject as being healthy or having an SRN. It was encouraging that many of the highest ranked OTUs were shared across these three models, (e.g. OTU 50, 426, 609, 822, 1239). The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU. However, this approach is only possible with linear models.

Second, to analyze non-linear models we interpreted the feature importance using permutation importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had a minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`. Effectively, the complexity of the communities was more fully represented in the better performing models [@sze_leveraging_2018; @wirbel_meta-analysis_2019]. At least in this case, it was not possible to distinguish between health and disease using a single OTU. Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU was associated with classifying a subject as being healthy or having an SRN.

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S4]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S4], 17 of the 20 top OTUs (e.g. OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. Although permutation analysis does not not allow us to determine the weight or the sign of the features, these results indicate that both methods are consistent in selecting the most important OTUs.

@@

There is a growing awareness that many human diseases and environmental processes are not driven by a single organism but are the product of multiple bacterial populations. Traditional statistical approaches are useful for identifying those cases where a single organism is associated with a process. In contrast, ML methods offer the ability to incorporate the structure of the microbial communities as a whole to classify them into different categories such as coming from a patient who is healthy or has SRNs. If it is possible to classify communities reliably, then ML methods also offer the ability to identify those microbial populations within the communities that are responsible for the classification. However, the application of ML in microbiome studies is still in its infancy and the field needs to develop a better understanding of different ML methods, their strengths and weaknesses, and how to implement them.

To address these needs, we developed a framework to train rigorous, transparent, and reproducible models. We benchmarked seven ML models and showed that we can create models that are inherently interpretable and easily trained without losing predictive performance. In terms of predictive performance, the random forest model had the best testing AUROC values compared to the other six models. However, the second-best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest is a complex ML model and could only be explained using post-hoc methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights. Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome. We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We repeated this data-split 100 times to measure the possible variation in predictive performance. During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the similar AUROC values from the cross-validation and testing.

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures which create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. For example if one OTU is highly correlated with another OTU, only one of these OTUs would have a large feature weight thus hiding the importance of its correlated OTU. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features. It is important to know the correlation structures of the data to avoid misinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. To identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.

In this study, we did not consider all possible modeling approaches. However, the principles highlighted throughout this study apply to other ML modeling tasks with microbiome data. For example, we did not evaluate multicategory classification methods to predict non-binary outcomes. We could have trained models to differentiate between people with normal colons and those with adenomas or carcinomas (k=3 categories). We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs. Furthermore, as number of categories to classify increase, more samples are required for each category to train a model. We also did not use regression-based analyses to predict a non-categorical outcome. We have previously used such an approach to train random forest models to predict fecal short-chain fatty acid concentrations based on microbiome data [@sze_fecal_2019]. Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks. Deep learning methods hold great promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which makes the deep learning models prone to overfitting. These methods are even more complex than random forest and XGBoost and are considered uninterpretable. There is great potential for applying ML approaches to microbiome data and we can choose which ML approach is best for the study.

@@

\newpage
\captionsetup{labelformat=empty}
\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
\small
\begin{tabular}{|l|l|l|l|l|}
\hline
